---
title: "Heart-Disease"
author: "Samah Abdelaal"
date: "`r Sys.Date()`"
output: html_document
---

## 9 - Restecg
   
Is restecg (resting ECG) different based on disease status? (Is there a significant difference in restecg between diseased and non-diseased groups?)

### Distribution of ca based on target groups(disease status)

```{r}
summary(new_heart$restecg)
```
  
```{r}
new_heart %>%
  summary_factorlist(dependent = "target_dummy", explanatory = "restecg", add_dependent_label = T, cont_cut = 1)
```
  
The prevalences of fixed and reversible defects in the disease group are **79.4** and **3.8%** respectively.  
     
```{r}
new_heart %>%
  ggplot(aes(x = thal, fill = thal)) +
  geom_bar() +
  labs(fill= "Thal") +
  facet_wrap(~ target_dummy) +
  theme_economist()
```
   
### Performing statistical tests  
   
Chi-square test  
  
```{r}
chisq.test(new_heart$thal, target_dummy , correct = F)
```
   
The p-value of Chi-square test is **significant(< 2.2e-16)**, so we **reject** the null hypothesis of no association between disease status and thal.  
