---
title: "Logistic Regression for Heart Disease UCI"
author: "Samah Abdelaal"
date: "11/24/2019"
output: word_document
---
Setup

```{r}
library(tidyverse)
library(knitr)
library(finalfit)
library(ggplot2)
library(ggthemes)
library(corrplot)
library(Greg)
```
Load data

```{r}
heart <- read_csv("Raw Data/heart.csv")
str(heart)
```

Data manipulation : setting up variables

```{r}
sex_dummy <- ifelse(heart$sex==1, "male" , "female")

cpasym <- ifelse(heart$cp==0, "1", "0")
cpatyp <- ifelse(heart$cp==1, "1", "0")
cpnon_ang <- ifelse(heart$cp==2, "1", "0")
cptyp <- ifelse(heart$cp==3, "1", "0")

fbs_dummy <- ifelse(heart$fbs==1, ">120", "<120")

restecg_notnorm <- ifelse(heart$restecg==0, "1", "0")
restecg_norm <- ifelse(heart$restecg==1, "1", "0")
restecg_abnor <- ifelse(heart$restecg==2, "1", "0")

exang_dummy <- ifelse(heart$exang==1, "yes", "no")

slope_down <- ifelse(heart$slope==0, "1", "0")
slope_flat <- ifelse(heart$slope==1, "1", "0")
slope_up <- ifelse(heart$slope==2, "1", "0")

thal_zero <- ifelse(heart$thal==0, "1", "0")
thal_nor <- ifelse(heart$thal==1, "1", "0")
thal_fixed <- ifelse(heart$thal==2, "1", "0")
thal_reves <- ifelse(heart$thal==3, "1", "0")

target_dummy <- ifelse(heart$target==1, "disease", "no disease")

heart$cp <- case_when(heart$cp==0 ~ "asymptomatic",
                      heart$cp==1 ~ "atypical angina",
                      heart$cp==2 ~ "non-anginal pain",
                      TRUE ~ "typical angina")

heart$restecg <- case_when(heart$restecg==0 ~ "LVH",
                           heart$restecg==1 ~ "normal",
                           TRUE ~ "ST")

heart$slope <- case_when(heart$slope==0 ~ "downsloping",
                         heart$slope==1 ~ "flat",
                         TRUE ~ "upsloping")

heart$thal <- case_when(heart$thal==1 ~ "reversible",
                        heart$thal==2 ~ "fixed",
                        TRUE ~ "normal")

new_heart <- cbind(heart, sex_dummy, cpasym, cpatyp, cpnon_ang, cptyp, fbs_dummy, restecg_notnorm, restecg_norm, restecg_abnor, exang_dummy, slope_down, slope_flat, slope_up, thal_zero, thal_nor, thal_fixed, thal_reves, target_dummy)

write_csv(new_heart, "Data/new_heart.csv")
```

```{r}
new_heart <- read.csv("Data/new_heart.csv")
attach(new_heart)
```

Data Summarization

```{r}
summary(new_heart)
```

, with sex composition of **207** males vs **96** females.  


# Analysis of Variables

## 1- Age

Is the age distribution different based on disease status? (Is there a significant difference in age between diseased and non-diseased groups?)

### Distribution of age based on target groups(disease status)

```{r}
summary(new_heart$age)
```

```{r}
new_heart %>%
  group_by(target_dummy) %>%
  summarise(count = n(), mean = mean(age), sd = sd(age), var = var(age))
```
  
**165** of the sample is diseased vs **138** with no disease.
Mean age in the diseased group is **52.6±9.55** vs **56.7±7.92** in the non-diseased group.  

```{r}
new_heart %>%
  ggplot(aes(x= age, fill= age)) +
  geom_histogram(binwidth = 1, color ="black", fill = "violetred2") +
  facet_wrap(~target_dummy)+
  theme_classic()
```
```{r}
boxplot(formula = age ~ target_dummy, data = new_heart)
```

### Checking assumptions of normality in age variable

```{r}
summary(new_heart$age)
```
  
   
The age of the sample ranges from **29** to **77** years old.The value of the age mean is slightly less than that of the age median, suggesting the distribution is left-skewed.   
    
Confirming with histogram:   
  
```{r}
hist(new_heart$age, col = "cadetblue")
```
  
```{r}
library(moments)
```
```{r}
skewness(new_heart$age)
```
  
Skewness value is negative(left-skewed)  
  
```{r}
kurtosis(new_heart$age)
```
  
Since kurtosis value is less than 3 (value of the normal distribution curve), The distribution is platykurtic.  
  
```{r}
shapiro.test(new_heart$age)
```
  
The p-value of Shapiro test is less than 0.05(0.006045), therefore the normality assumption is violated, age is **not normally distributed** and parametric tests cannot be applied.  
  
### Performing statistical tests  
  
Wilcoxon rank sum test  
  
```{r}
wilcox.test(new_heart$age ~ new_heart$target_dummy, mu=0, alternative = "two.sided", conf.level= 0.95, var.equal = F)
```
  
The p-value of Wilcoxon rank sum test is **highly significant(4.314e-05)**, so we **reject** the null hypothesis of no difference in age between diseased and not diseased.
  
## 2- Sex
  
Is the sex distribution different based on disease status? (Is there a significant difference in sex between diseased and non-diseased groups?)
  
### Distribution of age based on target groups(disease status)
  
```{r}
new_heart %>%
  summary_factorlist(dependent = "sex_dummy", explanatory = "target_dummy", add_dependent_label = T, cont_cut = 1)
```
   
The disease is more common in females than males(**74.7%** versus **44.3%** respectively).  
   
```{r}
new_heart %>%
  ggplot(aes(x = target_dummy, fill = target_dummy)) +
  geom_bar() +
  labs(fill= "Disease Status") +
  facet_wrap(~ sex_dummy) +
  theme_wsj()
```
   
### Performing statistical tests  
   
Chi-square test  
  
```{r}
chisq.test(target_dummy, sex_dummy, correct = F)
```
   
The p-value of Chi-square test is **highly significant(9.161e-07)**, so we **reject** the null hypothesis of no difference in sex between diseased and not diseased.  
   
## 3- FBS : fasting blood sugar
   
Is the disease status different based on fasting blood sugar(FBS)? (Is there a significant difference in disease status between high and low FBS groups?)  
  
```{r}
new_heart %>%
  summary_factorlist(dependent = "fbs_dummy", explanatory = "target_dummy", add_dependent_label = T, cont_cut = 1)
```
  
The disease prevalence **is 53.5%** in the high FBS group(>120 mg/dl) and **54.2%** in the low FBS group(<120 mg/dl).  
     
```{r}
new_heart %>%
  ggplot(aes(x = target_dummy, fill = target_dummy)) +
  geom_bar() +
  labs(fill= "Disease Status") +
  facet_wrap(~ fbs_dummy) +
  theme_wsj()
```
   
### Performing statistical tests  
   
Chi-square test  
  
```{r}
chisq.test(target_dummy, fbs_dummy, correct = F)
```
   
The p-value of Chi-square test is **not significant(0.9358)**, so we **fail to reject** the null hypothesis of no difference in disease status between high and low FBS groups.  
   
## 4- trestbps : Resting Blood Pressure RBP

Is resting blood pressure different based on disease status? (Is there a significant difference in resting blood pressure between diseased and non-diseased groups?)

### Distribution of RBP based on target groups(disease status)

```{r}
summary(new_heart$trestbps)
```

```{r}
new_heart %>%
  group_by(target_dummy) %>%
  summarise(mean = mean(trestbps), sd = sd(trestbps), var = var(trestbps))
```
  
Mean RBP in the diseased group is **129.2±16.4** vs **134.5±18.9** in the non-diseased group.  

```{r}
new_heart %>%
  ggplot(aes(x= trestbps, fill= trestbps)) +
  geom_histogram(binwidth = 3, color ="black", fill = "violetred2") +
  facet_wrap(~target_dummy)+
  theme_classic()
```

  
```{r}
boxplot(formula = trestbps ~ target_dummy, data = new_heart)
```
  
There are two notable outliers in the no disease group with resting blood pressure > 180 mm Hg   
  
### Checking assumptions of normality in RBP variable

```{r}
summary(new_heart$trestbps)
```
  
   
The RBP of the sample ranges from **94** to **200** mm Hg on admission to the hospital. The value of the mean is slightly higher than that of the median, suggesting the distribution is right-skewed.   
    
Confirming with histogram:   
  
```{r}
hist(new_heart$trestbps, col = "cadetblue2")
```
   
```{r}
library(moments)
```
```{r}
skewness(new_heart$trestbps)
```
  
Skewness value is positive(right-skewed)  
  
```{r}
kurtosis(new_heart$trestbps)
```
  
Since kurtosis value is more than 3 (value of the normal distribution curve), The distribution is leptokurtic.  
  
```{r}
shapiro.test(new_heart$trestbps)
```
  
The p-value of Shapiro test is less than 0.05(1.913e-06), therefore the normality assumption is violated, RBP is **not normally distributed** and parametric tests cannot be applied.  
  
### Performing statistical tests  
  
Wilcoxon rank sum test  
  
```{r}
wilcox.test(new_heart$trestbps ~ new_heart$target_dummy, mu=0, alternative = "two.sided", conf.level= 0.95, var.equal = F)
```
  
The p-value of Wilcoxon rank sum test is **significant(0.02918)**, so we **reject** the null hypothesis of no difference in RBP between diseased and not diseased.
      
## 5- chol : Serum Cholesterol

Is serum cholesterol different based on disease status? (Is there a significant difference in serum cholesterol between diseased and non-diseased groups?)

### Distribution of serum cholesterol based on target groups(disease status)

```{r}
summary(new_heart$chol)
```

```{r}
new_heart %>%
  group_by(target_dummy) %>%
  summarise(mean = mean(chol), sd = sd(chol), var = var(chol))
```
  
Mean cholesterol in the diseased group is **243.5±53.8** vs **251.5±49.7** in the non-diseased group.  
  
```{r}
new_heart %>%
  ggplot(aes(x= chol, fill= chol)) +
  geom_histogram(binwidth = 9, color ="black", fill = "violetred2") +
  facet_wrap(~target_dummy)+
  theme_classic()
```

  
```{r}
boxplot(formula = chol ~ target_dummy, data = new_heart)
```
  
There are four notable outliers with serum cholesterol > 400 mg/dl, three in the diseased group and one in the no disease group.   
  
### Checking assumptions of normality in chol variable

```{r}
summary(new_heart$chol)
```
  
   
The serum cholesterol of the sample ranges from **126** to **564** mg/dl. The value of the mean is slightly higher than that of the median, suggesting the distribution is right-skewed.   
    
Confirming with histogram:   
  
```{r}
hist(new_heart$chol, col = "cadetblue2")
```
   
```{r}
library(moments)
```
```{r}
skewness(new_heart$chol)
```
  
Skewness value is positive(right-skewed)  
  
```{r}
kurtosis(new_heart$chol)
```
  
Since kurtosis value is more than 3 (value of the normal distribution curve), The distribution is leptokurtic.  
  
```{r}
shapiro.test(new_heart$chol)
```
  
The p-value of Shapiro test is less than 0.05(8.986e-09), therefore the normality assumption is violated, serum cholesterol is **not normally distributed** and parametric tests cannot be applied.  
  
### Performing statistical tests  
  
Wilcoxon rank sum test  
  
```{r}
wilcox.test(new_heart$chol ~ new_heart$target_dummy, mu=0, alternative = "two.sided", conf.level= 0.95, var.equal = F)
```
  
The p-value of Wilcoxon rank sum test is **not significant(0.05563)**, so we **reject** the null hypothesis of no difference in serum cholesterol between diseased and not diseased.
            
## 6- thalach : Maximum Heart Rate Achieved

Is the maximum heart rate different based on disease status? (Is there a significant difference in maximum heart rate between diseased and non-diseased groups?)

### Distribution of maximum heart rate based on target groups(disease status)

```{r}
summary(new_heart$thalach)
```

```{r}
new_heart %>%
  group_by(target_dummy) %>%
  summarise(mean = mean(thalach), sd = sd(thalach), var = var(thalach))
```
  
Mean maximum heart rate in the diseased group is **158.6±19.04** vs **138.9±22.72** in the non-diseased group.  

```{r}
new_heart %>%
  ggplot(aes(x= thalach, fill= thalach)) +
  geom_histogram(binwidth = 5, color ="black", fill = "violetred2") +
  facet_wrap(~target_dummy)+
  theme_classic()
```
```{r}
boxplot(formula = thalach ~ target_dummy, data = new_heart)
```

### Checking assumptions of normality in thalach variable

```{r}
summary(new_heart$thalach)
```
  
   
The maximum heart rate of the sample ranges from **71** to **202** bpm. The value of the mean is slightly less than that of the median, suggesting the distribution is left-skewed.   
    
Confirming with histogram:   
  
```{r}
hist(new_heart$thalach, col = "coral3")
```
  
```{r}
library(moments)
```
```{r}
skewness(new_heart$thalach)
```
  
Skewness value is negative(left-skewed)  
  
```{r}
kurtosis(new_heart$thalach)
```
  
Since kurtosis value is close to 3 (value of the normal distribution curve), The distribution is mesokurtic.  
  
```{r}
shapiro.test(new_heart$thalach)
```
  
The p-value of Shapiro test is less than 0.05(0.0001073), therefore the normality assumption is violated, maximum heart rate is **not normally distributed** and parametric tests cannot be applied.  
  
### Performing statistical tests  
  
Wilcoxon rank sum test  
  
```{r}
wilcox.test(new_heart$thalach ~ new_heart$target_dummy, mu=0, alternative = "two.sided", conf.level= 0.95, var.equal = F)
```
  
The p-value of Wilcoxon rank sum test is **highly significant(1.251e-13)**, so we **reject** the null hypothesis of no difference in maximum heart rate between diseased and not diseased.
   
## 7- Oldpeak : (ST depression induced by exercise relative to rest)

Is oldpeak(ST depression induced by exercise relative to rest) different based on disease status? (Is there a significant difference in oldpeak between diseased and non-diseased groups?)

### Distribution of oldpeak based on target groups(disease status)

```{r}
summary(new_heart$oldpeak)
```

```{r}
new_heart %>%
  group_by(target_dummy) %>%
  summarise(mean = mean(oldpeak), sd = sd(oldpeak), var = var(oldpeak))
```
  
Mean oldpeak in the diseased group is **0.6±0.79** vs **1.6±1.3** in the non-diseased group.  

```{r}
new_heart %>%
  ggplot(aes(x= oldpeak, fill= oldpeak)) +
  geom_histogram(binwidth = 1, color ="black", fill = "violetred2") +
  facet_wrap(~target_dummy)+
  theme_classic()
```
```{r}
boxplot(formula = oldpeak ~ target_dummy, data = new_heart)
```
  
There are two notable outliers with ST depression > 5 in the no disease group.  
  
### Checking assumptions of normality in oldpeak variable

```{r}
summary(new_heart$oldpeak)
```
  
   
The oldpeak of the sample ranges from **0** to **6.2**.The value of the mean is higher than that of the median, suggesting the distribution is right-skewed.   
    
Confirming with histogram:   
  
```{r}
hist(new_heart$oldpeak, col = "salmon")
```
  
```{r}
library(moments)
```
```{r}
skewness(new_heart$oldpeak)
```
  
Skewness value is positive(right-skewed)  
  
```{r}
kurtosis(new_heart$oldpeak)
```
  
Since kurtosis value is more than 3 (value of the normal distribution curve), The distribution is leptokurtic.  
  
```{r}
shapiro.test(new_heart$oldpeak)
```
  
The p-value of Shapiro test is less than 0.05(2.501e-16), therefore the normality assumption is violated, oldpeak is **not normally distributed** and parametric tests cannot be applied.  
  
### Performing statistical tests  
  
Wilcoxon rank sum test  
  
```{r}
wilcox.test(new_heart$oldpeak ~ new_heart$target_dummy, mu=0, alternative = "two.sided", conf.level= 0.95, var.equal = F)
```
  
The p-value of Wilcoxon rank sum test is **highly significant(7.574e-13)**, so we **reject** the null hypothesis of no difference in oldpeak between diseased and not diseased.
  
   
## 8- ca : number of major vessels (0-3) colored by flourosopy
  
Is ca : number of major vessels (0-3) colored by flourosopy different based on disease status? (Is there a significant difference in oldpeak between diseased and non-diseased groups?)
  
### Distribution of ca based on target groups(disease status)
  
It does not make much sense to deal with ca as a numerical variable since it can only have 4 values (0,1,2,3) and no decimals, so it will be analyzed as a factor.
  
```{r}
new_heart$ca <- as.factor(new_heart$ca)
summary(new_heart$ca)
```
  

```{r}
new_heart %>%
  summary_factorlist(dependent = "target_dummy", explanatory = "ca", add_dependent_label = T, cont_cut = 1)
```
  
The prevalence of having zero, one, two, and three vessels colored by flouroscopy in the disease group is **80.6**, **13.1**, **4.4**, and **1.9%** respectively.  
     
```{r}
new_heart %>%
  ggplot(aes(x = ca, fill = ca)) +
  geom_bar() +
  labs(fill= "number of major vessels colored by flourosopy") +
  facet_wrap(~ target_dummy) +
  theme_economist()
```
   
### Performing statistical tests  
   
Chi-square test is incorrect in this case, so Fisher's exact test will be used instead.
  
```{r}
fisher.test(new_heart$ca, target_dummy)
```
   
The p-value of Fisher's exact test is **significant(< 2.2e-16)**, so we **reject** the null hypothesis of no association between disease status and number of major vessels colored by flouroscopy.  
  
## 9 - Restecg
   
Is restecg (resting ECG) different based on disease status? (Is there a significant difference in restecg between diseased and non-diseased groups?)

### Distribution of restecg based on target groups(disease status)

```{r}
summary(new_heart$restecg)
```
  
```{r}
new_heart %>%
  summary_factorlist(dependent = "target_dummy", explanatory = "restecg", add_dependent_label = T, cont_cut = 1)
```
  
The prevalence of probable or definite left ventricular hypertrophy (LVH) is **41.9** and in the disease group versus **57.4%** in the no disease group.  
  
The prevalence of ST-T wave abnormality is **0.6** in the disease group versus **2.2** in the no disease group.  
  
```{r}
new_heart %>%
  ggplot(aes(x = restecg, fill = restecg)) +
  geom_bar() +
  labs(fill= "Resting ECG") +
  facet_wrap(~ target_dummy) +
  theme_economist()
```
   
### Performing statistical tests  
   
Chi-square test is incorrect in this case, so Fisher's exact test will be used instead.   
```{r}
fisher.test(new_heart$restecg, target_dummy)
```
   
The p-value of Fisher's exact test is **significant(0.005866)**, so we **reject** the null hypothesis of no association between disease status and resting ECG.  

## 10 - Slope
   
Is slope (the slope of the peak exercise ST segment) different based on disease status? (Is there a significant difference in slope between diseased and non-diseased groups?)

### Distribution of slope based on target groups(disease status)

```{r}
summary(new_heart$slope)
```
  
```{r}
new_heart %>%
  summary_factorlist(dependent = "target_dummy", explanatory = "slope", add_dependent_label = T, cont_cut = 1)
```
  
The prevalence of upsloping versus downsloping ST segment in the disease group is **64.4** versus **5.6%**, and in the no disease group is **25.7** versus **8.8**.  
  
```{r}
new_heart %>%
  ggplot(aes(x = slope, fill = slope)) +
  geom_bar() +
  labs(fill= "Slope of peak exercise ST segment") +
  facet_wrap(~ target_dummy) +
  theme_economist()
```
   
### Performing statistical tests  
   
Chi-square test   

```{r}
chisq.test(new_heart$slope, target_dummy, correct = F)
```
   
The p-value of Chi-square test is **significant(2.116e-10)**, so we **reject** the null hypothesis of no association between disease status and slope.  
  
## 11 - Thal
   
Is thal different based on disease status? (Is there a significant difference in thal between diseased and non-diseased groups?)

### Distribution of ca based on target groups(disease status)

```{r}
summary(new_heart$thal)
```
  
```{r}
new_heart %>%
  summary_factorlist(dependent = "target_dummy", explanatory = "thal", add_dependent_label = T, cont_cut = 1)
```
  
The prevalences of fixed and reversible defects in the disease group are **79.4** and **3.8%** respectively.  
     
```{r}
new_heart %>%
  ggplot(aes(x = thal, fill = thal)) +
  geom_bar() +
  labs(fill= "Thal") +
  facet_wrap(~ target_dummy) +
  theme_economist()
```
   
### Performing statistical tests  
   
Chi-square test  
  
```{r}
chisq.test(new_heart$thal, target_dummy , correct = F)
```
   
The p-value of Chi-square test is **significant(< 2.2e-16)**, so we **reject** the null hypothesis of no association between disease status and thal.  
  
## 12 - Cp
   
Is cp (chest pain type) different based on disease status? (Is there a significant difference in chest pain type between diseased and non-diseased groups?)

### Distribution of cp based on target groups(disease status)

```{r}
summary(new_heart$cp)
```
  
```{r}
new_heart %>%
  summary_factorlist(dependent = "target_dummy", explanatory = "cp", add_dependent_label = T, cont_cut = 1)
```
  
The prevalence of typical angina, atypical angina and non-angina in the disease group is **10%**, **25%** and **40.6%** in the disease group versus **5.1%**, **6.6%** and **13.2%** in the no disease group.  
  
```{r}
new_heart %>%
  ggplot(aes(x = cp, fill = cp)) +
  geom_bar() +
  labs(fill= "Chest pain type") +
  facet_wrap(~ target_dummy) +
  theme_economist() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 0.6))
```
   
### Performing statistical tests  
   
Chi-square test    

```{r}
chisq.test(new_heart$cp, target_dummy)
```
   
The p-value of Chi-square test is **significant(< 2.2e-16)**, so we **reject** the null hypothesis of no association between disease status and chest pain type.  

## 13 - Exang
   
Is exang (exercise-induced angina) different based on disease status? (Is there a significant difference in exercise-induced angina between diseased and non-diseased groups?)

### Distribution of exang based on target groups(disease status)

```{r}
summary(new_heart$exang_dummy)
```
  
```{r}
new_heart %>%
  summary_factorlist(dependent = "target_dummy", explanatory = "exang_dummy", add_dependent_label = T, cont_cut = 1)
```
  
The prevalence of exercise-induced angina in the disease group is **14.4** versus **54.4%** in the no disease group.   
   
```{r}
new_heart %>%
  ggplot(aes(x = exang_dummy, fill = exang_dummy)) +
  geom_bar() +
  labs(fill= "Exercise induced angina") +
  facet_wrap(~ target_dummy) +
  theme_economist()
```
   
### Performing statistical tests  
   
Chi-square test   

```{r}
chisq.test(new_heart$exang_dummy, target_dummy, correct = F)
```
   
The p-value of Chi-square test is **significant(2.604e-13)**, so we **reject** the null hypothesis of no association between disease status and exercise-induced angina.  
